\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-26 coqimpl.tex
Rewrite global reasoning section.

The simplest idea is to allow partiality---i.e., returning
\mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.
We call this \textit{propositional partial recursor}.
For example, the propositional partial recursor for \mintinline{Coq}{B.b} is 
\begin{minted}[escapeinside=@@]{Coq}
b_prec : forall P : b@\YZ{b or self_B.b?}@ -> Set, P tt -> P ff
    -> forall x : b, option (P x).
\end{minted}
with appropriate computational axioms. The partiality doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec").

@@

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> R) -> R)), .., D -> R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using eliminator\YZ{eliminator of what?} and "prec"
  (both acting like identity) such that
  \begin{minted}{Coq}
    forall c : C, linv (inj c) = some c
  \end{minted}
\end{theorem}

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
However, to verify and prove the recursor and partial recursor and their
computational rules (as in \ref{langdesign-req4}), our plugin has to
compile the extensible family into a non-extensible Coq Module and define
the recursors by instantiating the content of Coq module with concrete
inductive types.

This style of ``Global Reasoning'' says that we can prove the
corresponding theorem only when we know the ``full picture'', in particular,
all the constructors of an inductive type. We provide "Closing Fact" as
a command to carry out this global reasoning\YZ{Are recursors and partial recursors automatically generated and proved by the plugin, or must they be asserted and proved by the programmer?}---we can attach to "Closing
Fact" either (proof) terms or Coq's proof script, then only when
compilation to module happens, is the term/proof script
type-checked to see if a "Closing Fact" statement can be proved. Note
that, "Closing Fact" cannot immediately check the correctness of the
term/proof script because this verification only happens when we
finished defining a family, and thus a bit harder to handle.

This "Closing Fact" can ease us from the necessity to let plugin
generate proof for recursors and partial recursors and their
computational rules (as in \ref{langdesign-req4}), and thus making
plugin development easier.\YZ{This reads like a bad excuse: the programmer does not care about if the plugin development is easy. Question is if it makes the programmer's life easier.}
For example, the computation rule for a
concrete recursor like "B.neg" and partial recursors in
\cref{fig:plugin-example2} can be directly asserted by using "Closing
Fact" and their proofs verified later during compilation.\YZ{What is the alternative if not using "Closing Fact"?}
This ``Global Reasoning'' will be again useful in later examples.\YZ{Still not entirely clear to me what is gained and what is lost by using "Closing Fact".}

>>>>

The simplest idea is to allow partiality---i.e., returning
\mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.
We call this \textit{propositional partial recursor}.
For example, the propositional partial recursor for \mintinline{Coq}{B.b} is 
\begin{minted}[escapeinside=@@]{Coq}
b_prec : forall P : b@\YZ{b or self_B.b?}\EDJreply{it is mainly used as self_B.b, but to make option(P x) at the end doesn't require to be self_B.b so it is kind of orthognal}@ -> Set, P tt -> P ff
    -> forall x : b, option (P x).
\end{minted}
with appropriate computational axioms. The partiality doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec").

@@

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> R) -> R)), .., D -> R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of "C"\YZ{eliminator of what?}\EDJreply{"D" is not inductive type so it doesn't have eliminator} and "prec" of "D"
  (both acting like identity) such that
  \begin{minted}{Coq}
    forall c : C, linv (inj c) = some c
  \end{minted}
\end{theorem}

@@

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
Unfortunately, there are cases the users just want global reasoning instead of an extensible proof. For example, in STLC, we should expect`` \mintinline{Coq}{~ value (tm_app x y)}'' holds in all possible extension of STLC (i.e. application is never a terminal value). We know the correct way of proving this proposition \mintinline{Coq}{Field app_not_value : value (tm_app x y) -> False.} is to use "FRecursor" to inductively reason "value (tm_app x y)". However, this means that everytime we extend "value" with new terminal value, we will need to extend corresponding inductive cases for "app_not_value". We can imagine the newly extended proof to be boring case analysis that the newly extended terminal value is not "tm_app". What's worse, similar scenario can happen on other statement like \mintinline{Coq}{~ value (tm_if cond x y)}. Everything here is extended correctly but we end up having boilerplate code, \textbf{semantically}.

Our proposal is to allow global reasoning, which will look like a constraint inside the family. We provide "Closing Fact" as a command to carry out this global reasoning, attached with a proof script and proof term. This "Closing Fact" will look like an assertion during the construction of the surrounding family, and this assertion will only be verified when the whole family is closed and compiled into a module, with the help of the attached proof script and proof terms.

For example,

The downside is that "Closing Fact" cannot immediately check the correctness of the
attached proof term/proof script because this verification only happens when we
finished defining a family, and thus a bit harder to handle.

% However, to verify and prove the recursor and partial recursor and their
% computational rules (as in \ref{langdesign-req4}), our plugin has to
% compile the extensible family into a non-extensible Coq Module and define
% the recursors by instantiating the content of Coq module with concrete
% inductive types. We provide "Closing Fact" as a command to carry out this global reasoning,

% This style of ``Global Reasoning'' says that we can prove the
% corresponding theorem only when we know the ``full picture'', in particular,
% all the constructors of an inductive type. 
% We provide "Closing Fact" as
% a command to carry out this global reasoning
\YZ{Are recursors and partial recursors automatically generated and proved by the plugin, or must they be asserted and proved by the programmer?}\EDJreply{(1) Closing Fact is not only used to generate recursor and partial recursor. I think the intro of this section is just giving people wrong impression. So I rewrite the whole sectio now. (2)  Recursors and Partial recursors are possible to be automatically generated but my plugin didn't implement it. Current stage is that user only need to assert them. Nobody has to prove it (plugin will prove it). Here when I say user has to provide proof script, it means for arbitrary proposition if the user want to carry out global reasoning.}
% ---we can attach to "Closing
% Fact" either (proof) terms or Coq's proof script, then only when
% compilation to module happens, is the term/proof script
% type-checked to see if a "Closing Fact" statement can be proved. Note
% that, "Closing Fact" cannot immediately check the correctness of the
% term/proof script because this verification only happens when we
% finished defining a family, and thus a bit harder to handle.

% This "Closing Fact" also can ease us from the necessity to let plugin
% generate proof for recursors and partial recursors and their
% computational rules (as in \ref{langdesign-req4}), and thus making
% plugin development easier.
\YZ{This reads like a bad excuse: the programmer does not care about if the plugin development is easy. Question is if it makes the programmer's life easier.}\EDJreply{You are right. It does sound like a bad excuse. I rewrite the whole section now. Closing Fact is initially used when the user wants global reasoning instead of extensible proof by induction, for example, the inversion lemma. \\ It just happens that they can be used by me so that I can postpone generation of these rule plugin (which is hard for me to do). This paragraph is just me being honest that my plugin is not mature enough to generate everything.}
% For example, the computation rule for a
% concrete recursor like "B.neg" and partial recursors in
% \cref{fig:plugin-example2} can be directly asserted by using "Closing
% Fact" and their proofs verified later during compilation.
\YZ{What is the alternative if not using "Closing Fact"?}\EDJreply{To be a legit plugin, my plugin should generate them...}
% This ``Global Reasoning'' will be again useful in later examples.\YZ{Still not entirely clear to me what is gained and what is lost by using "Closing Fact".}
